---
title: "Blantyre Air Quality Modelling"
output: html_notebook
---

Load required libraries

```{r}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)
library(sf)
library(mgcv)
library(priorsense)
library(osmdata)
library(nngeo) 
```


Import the modelling dataset - cleaned and prepped in the `2025-05-12_aq_clean.Rmd` script

```{r}

aq_model_data <- read_rds("input_data/aq_model_data.rds")

#also load the scale clusters
load("input_data/scale_72_clusters.rda") #SCALE clusters

#and the grid data
mw_100m_cropped <- read_rds("input_data/mw_100m_cropped.rds")
mw_100m_grid_sf <- read_rds("input_data/mw_100m_grid_sf.rds")


```


### Model 1:


"We fitted a spatially smooth regression model for log-transformed PM2.5 concentrations using a Gaussian process (GP) smooth over spatial coordinates (x, y) to capture flexible spatial trends. Seasonal variation was modelled using a Fourier series expansion of day-of-year up to the 3rd harmonic (i.e., sine and cosine terms for annual, semi-annual, and tri-annual cycles) to account for complex seasonal patterns. The model was fitted in a Bayesian framework using brms with a Gaussian likelihood and the cmdstanr backend."

Note, priors are still causing issues for me, so we will comment out now, and use the defauly stan weakly informative priors

Priors

```{r, fig.width=12, fig.height=12}

priors <- c(
  prior(normal(3.4, 1), class = "Intercept"),
  prior(normal(0,1), class = "b"),
  prior(exponential(1), class = "sigma"),
  prior(exponential(1), class = "sdgp"),
  prior(normal(0, 1), class = "lscale", lb = 0)
)

```

Fit prior only model

```{r, fig.width=12, fig.height=12}

m1_prior <- brm(
    formula = log_pm2_5 ~ gp(x, y, k=15) +
      sin_doy1 + cos_doy1 +
      sin_doy2 + cos_doy2 +
      sin_doy3 + cos_doy3,
    data = aq_model_data,
    family = gaussian(),
    prior = priors,
    sample_prior = "only",
    chains = 4, cores = 4,
    backend = "cmdstanr"
  )


summary(m1_prior)
plot(m1_prior)

```

Now model with data.


```{r, fig.width=12, fig.height=12}
m1 <- brm(
    formula = log_pm2_5 ~ gp(x, y, k=15)  +
    sin_doy1 + cos_doy1 +
    sin_doy2 + cos_doy2 +
    sin_doy3 + cos_doy3,
    data = aq_model_data,
    family = gaussian(),
    prior = priors,
    chains = 4, cores = 4,
    backend = "cmdstanr"
  )

summary(m1)
conditional_effects(m1)
pp_check(m1)
plot(m1)

```

Compare prior-only model to model with data using `priorsense` package:

```{r, fig.width=12, fig.height=12}

m1_draws <- as_draws_df(m1)

# Extract prior and posterior draws
powerscale_sensitivity(m1, variable = c("b_Intercept", 
                                        "b_sin_doy1", "b_cos_doy1",
                                        "b_sin_doy2", "b_cos_doy2",
                                        "b_sin_doy3", "b_cos_doy3",
                                        "sdgp_gpxy", "sigma", "Intercept"))

powerscale_plot_dens(m1, variable = c("b_Intercept", 
                                        "b_sin_doy1", "b_cos_doy1",
                                        "b_sin_doy2", "b_cos_doy2",
                                        "b_sin_doy3", "b_cos_doy3",
                                        "sdgp_gpxy", "sigma", "Intercept"))
```

Predictions by space and time. Here, although we only have data within clusters, we will predict for cells outside of clusters based on covariates.
We will also predict for the month for whcih we do not have data.

```{r, fig.width=12, fig.height=12}

#set_up prediction date frame
nd_m1 <- st_as_sf(mw_100m_cropped)  %>%
  mutate(centroid = st_centroid(geometry)) %>%
  mutate(x = st_coordinates(centroid)[, 1],
         y = st_coordinates(centroid)[, 2]) %>%
  select(-centroid)  %>%
  st_drop_geometry() %>%
  rename(pop_density = 1) %>%
  mutate(pop_density_km2 = pop_density / 0.01)

#get the first day of each month for prediction
first_days <- ymd(paste0("2020-", sprintf("%02d", 1:12), "-01"))
first_day_doy <- yday(first_days)

#calculate Fourier terms
first_day_seasonality <- tibble(
  mean_doy = first_day_doy,
  sin_doy1 = sin(2 * pi * mean_doy / 365.25),
  cos_doy1 = cos(2 * pi * mean_doy / 365.25),
  sin_doy2 = sin(4 * pi * mean_doy / 365.25),
  cos_doy2 = cos(4 * pi * mean_doy / 365.25),
  sin_doy3 = sin(6 * pi * mean_doy / 365.25),
  cos_doy3 = cos(6 * pi * mean_doy / 365.25),
  date = first_days
)

#expand grid for prediction
nd_m1 <- nd_m1 %>%
  crossing(first_day_seasonality)

#take posterior draws - note storing as an rvars object for efficiency
#note can only manage this without crashing computer!
#to address later, and improve efficieny for more draws
m1_draws <- add_epred_rvars(object = m1,
                            newdata = nd_m1,
                            ndraws = 500) 

#now extract the summary mean and sd for each month-grid cell combination
m1_epred_array <- posterior::draws_of(m1_draws$.epred)

#compute posterior means per location
m1_draws$.epred_mean <- colMeans(m1_epred_array)

#compute posterior sds per location
m1_draws$.epred_sd <- apply(m1_epred_array, 2, sd)

m1_sum <- m1_draws %>% 
  select(x, y, date, .epred_mean, .epred_sd)


#plot predictions
m1_sum %>%  
  mutate(date = month(date, label=TRUE)) %>%
  ggplot() +
  geom_tile(aes(x = x, y=y, fill=.epred_mean)) +
  geom_sf(data = scale_72_clusters, colour="grey78", fill=NA) +
  scale_fill_viridis_c() +
  facet_wrap(date~.) +
  labs(title = "Log(PM2.5) exposure",
       x="",
       y="") +
  theme_ggdist() +
  theme(panel.background = element_rect(fill=NA, colour="grey78"),
        axis.text.x = element_text(angle = 45, hjust=1),
        strip.background = element_rect(colour = "grey78"))

```

Now just draw 50 random coordinates, and predict by time to see whether we captured the trend.

```{r, fig.width=12, fig.height=12}
#sample coordinate points from the household dataset
#we will fix the small number sampled later!
set.seed(123) 
sampled_points_m1 <- aq_model_data %>%
  sample_n(50) %>%
  select(x, y)

#generate DOY 0–365 and calculate Fourier terms
doy_grid <- tibble(doy = 0:365) %>%
  mutate(
    sin_doy1 = sin(2 * pi * doy / 365),
    cos_doy1 = cos(2 * pi * doy / 365),
    sin_doy2 = sin(4 * pi * doy / 365),
    cos_doy2 = cos(4 * pi * doy / 365),
    sin_doy3 = sin(6 * pi * doy / 365),
    cos_doy3 = cos(6 * pi * doy / 365)
  )

#expand grid across sampled locations
prediction_df_m1 <- sampled_points_m1 %>%
  crossing(doy_grid)

#add predictions
preds_m1 <- add_epred_draws(object = m1, newdata = prediction_df_m1)

#summarise
preds_m1_sum <- preds_m1 %>%
  ungroup() %>%
  group_by(doy) %>%
  summarise(.epred_mean = mean(.epred),
            .lower = quantile(.epred, probs=0.025),
            .upper = quantile(.epred, probs = 0.975))

#GEt the observed data
observed_data <- aq_model_data %>%
  select(mean_doy, log_pm2_5) 

#plot
preds_m1_sum %>%
  ggplot() +
  geom_ribbon(aes(x=doy, ymin=.lower, ymax = .upper), fill="steelblue", alpha=0.3) +
  geom_line(aes(x=doy, y=.epred_mean)) +
  geom_jitter(data = observed_data, aes(x = mean_doy, y = log_pm2_5),
              color = "darkred", alpha = 0.6, width = 0.5, height = 0.0, size = 1.2) +
  labs(title = "Model-estimated log(PM2.5) with empirical measurements",
       subtitle = "Model predictions with 95% CrI and observed data points",
       x = "Day of year",
       y = "log(PM2.5)",
       caption = "Modelled estimates restricted to within clusters") +
  theme_ggdist() +
  theme(panel.background = element_rect(colour = "grey78"),
        legend.position = "none")

```

### Model 2: With covariates

Now we include grid level covariates of distance to the road, population density, and building footprint percent, along witht the mean temp and huidity on the day of measurement (from the purple air monitors)

Again, priors to be fixed later

```{r, fig.width=12, fig.height=12}
priors <- c(
  prior(normal(3.4, 1), class = "Intercept"),
  prior(normal(0,1), class = "b"),
  prior(exponential(1), class = "sigma"),
  prior(exponential(1), class = "sdgp"),
  prior(normal(0, 1), class = "lscale", lb = 0)
)


m2 <- brm(
    formula = log_pm2_5 ~ gp(x, y, k=15)  +
      s(mean_temp_c, k=5) + 
      s(mean_current_humidity, k=5) +
      s(pop_density_km2, k=5) +
      s(building_coverage_pct, k=5) +
      s(dist_to_road_m, k=5) +
      sin_doy1 + cos_doy1 +
      sin_doy2 + cos_doy2 +
      sin_doy3 + cos_doy3,
    data = aq_model_data,
    family = gaussian(),
    prior = priors,
    chains = 4, cores = 4,
    backend = "cmdstanr"
  )

summary(m2)
conditional_effects(m2)
pp_check(m2)
plot(m2)


```

```{r, fig.width=12, fig.height=12}

m2_draws <- as_draws_df(m2)

# Extract prior and posterior draws
powerscale_sensitivity(m2, variable = c("b_Intercept", "b_sin_doy1", "b_cos_doy1", "b_sin_doy2", "b_cos_doy2", "b_sin_doy3", "b_cos_doy3", "bs_smean_temp_c_1", "bs_smean_current_humidity_1", "bs_spop_density_km2_1", "bs_sbuilding_coverage_pct_1", "bs_sdist_to_road_m_1",         
"sds_smean_temp_c_1", "sds_smean_current_humidity_1", "sds_spop_density_km2_1", "sds_sbuilding_coverage_pct_1", 
"sds_sdist_to_road_m_1", "sdgp_gpxy", "lscale_gpxy", "sigma", "Intercept", "s_smean_temp_c_1[1]", "s_smean_temp_c_1[2]", "s_smean_temp_c_1[3]",         "s_smean_current_humidity_1[1]",  "s_smean_current_humidity_1[2]",  "s_smean_current_humidity_1[3]",  "s_spop_density_km2_1[1]",       
 "s_spop_density_km2_1[2]",        "s_spop_density_km2_1[3]",        "s_sbuilding_coverage_pct_1[1]",  "s_sbuilding_coverage_pct_1[2]", 
 "s_sbuilding_coverage_pct_1[3]",  "s_sdist_to_road_m_1[1]", "s_sdist_to_road_m_1[2]", "s_sdist_to_road_m_1[3]"))

powerscale_plot_dens(m2, variable = c("b_Intercept", "b_sin_doy1", "b_cos_doy1", "b_sin_doy2", "b_cos_doy2", "b_sin_doy3", "b_cos_doy3", "bs_smean_temp_c_1", "bs_smean_current_humidity_1", "bs_spop_density_km2_1", "bs_sbuilding_coverage_pct_1", "bs_sdist_to_road_m_1",         
"sds_smean_temp_c_1", "sds_smean_current_humidity_1", "sds_spop_density_km2_1", "sds_sbuilding_coverage_pct_1", 
"sds_sdist_to_road_m_1", "sdgp_gpxy", "lscale_gpxy", "sigma", "Intercept", "s_smean_temp_c_1[1]", "s_smean_temp_c_1[2]", "s_smean_temp_c_1[3]",         "s_smean_current_humidity_1[1]",  "s_smean_current_humidity_1[2]",  "s_smean_current_humidity_1[3]",  "s_spop_density_km2_1[1]",       
 "s_spop_density_km2_1[2]",        "s_spop_density_km2_1[3]",        "s_sbuilding_coverage_pct_1[1]",  "s_sbuilding_coverage_pct_1[2]", 
 "s_sbuilding_coverage_pct_1[3]",  "s_sdist_to_road_m_1[1]", "s_sdist_to_road_m_1[2]", "s_sdist_to_road_m_1[3]"))
```

Predictions by space and time

```{r, fig.width=12, fig.height=12}

#first we need a prediction dataframe with all of the covariate data

#extract month from date variable
aq_model_data <- aq_model_data  %>%
  mutate(month = lubridate::month(as.Date(mean_doy, origin = "2019-12-31")))

#calculate monthly means
monthly_means <- aq_model_data %>%
  group_by(month) %>%
  summarise(
    mean_temp_c = mean(mean_temp_c, na.rm = TRUE),
    mean_current_humidity = mean(mean_current_humidity, na.rm = TRUE)
  )

#No data collection in April - here we will just interpolate
#can do better later...
march_temp <- monthly_means %>% filter(month == 3) %>% pull(mean_temp_c)
may_temp <- monthly_means %>% filter(month == 5) %>% pull(mean_temp_c)
april_temp <- (march_temp + may_temp) / 2

march_humidity <- monthly_means %>% filter(month == 3) %>% pull(mean_current_humidity)
may_humidity <- monthly_means %>% filter(month == 5) %>% pull(mean_current_humidity)
april_humidity <- (march_humidity + may_humidity) / 2

#add these into the prediction dataframe
monthly_means <- monthly_means %>%
  add_row(month=4, 
          mean_temp_c = april_temp,
          mean_current_humidity = april_humidity) %>%
  arrange(month)

#set_up prediction date frame
nd_m2 <- mw_100m_grid_sf %>%
  sf::st_sf() %>%
  mutate(x = st_coordinates(st_centroid(.))[, 1],
         y = st_coordinates(st_centroid(.))[, 2]) %>%
  st_drop_geometry() %>%
  rename(pop_density = 1) %>%
  mutate(pop_density_km2 = pop_density / 0.01)

#get the first day of each month for prediction
first_days <- ymd(paste0("2020-", sprintf("%02d", 1:12), "-01"))
first_day_doy <- yday(first_days)

#calculate Fourier terms
first_day_seasonality <- tibble(
  mean_doy = first_day_doy,
  sin_doy1 = sin(2 * pi * mean_doy / 365.25),
  cos_doy1 = cos(2 * pi * mean_doy / 365.25),
  sin_doy2 = sin(4 * pi * mean_doy / 365.25),
  cos_doy2 = cos(4 * pi * mean_doy / 365.25),
  sin_doy3 = sin(6 * pi * mean_doy / 365.25),
  cos_doy3 = cos(6 * pi * mean_doy / 365.25),
  date = first_days
) %>%
  mutate(month = lubridate::month(as.Date(mean_doy, origin = "2019-12-31"))) %>%
  left_join(monthly_means)

nd_m2 <- nd_m2 %>%
  crossing(first_day_seasonality) %>%
  mutate(building_coverage_pct = building_coverage_pct/100)

#take posterior draws - note storing as an rvars object for efficiency
m2_draws <- add_epred_rvars(object = m2,
                            newdata = nd_m2,
                            ndraws = 200) #note can only manage this without crashing computer!

#now extract the summary mean and sd for each month-grid cell combination
m2_epred_array <- posterior::draws_of(m2_draws$.epred)

#compute posterior means per location
m2_draws$.epred_mean <- colMeans(m2_epred_array)

#compute posterior sd per location
m2_draws$.epred_sd <- apply(m2_epred_array, 2, sd)

m2_sum <- m2_draws %>% 
  select(x, y, date, .epred_mean, .epred_sd)


#plot predictions
m2_sum %>%  
  mutate(date = month(date, label = TRUE)) %>%
  mutate(.epred_mean = exp(.epred_mean)) %>%
  ggplot() +
  geom_tile(aes(x = x, y=y, fill=.epred_mean)) +
  geom_sf(data = scale_72_clusters, colour="grey78", fill=NA) +
  #geom_sf(data = main_roads_cropped, colour="white") + #roads looking a bit messy when plotted - work on this.
  scale_fill_scico(palette = "acton",
    trans = scales::log_trans(base = 10), name = "Predicted\nPM2.5") +
  facet_wrap(date~.) +
  labs(title = "Log(PM2.5) exposure",
       x="",
       y="",
       caption = "White boundaries are SCALE study clusters") +
  theme_ggdist() +
  theme(panel.background = element_rect(fill=NA, colour="grey78"),
        strip.background = element_rect(colour = "grey78"),
        axis.text.x = element_text(angle=90, hjust=1),
        legend.position = "bottom")

#plot sd
m2_sum %>%  
  mutate(date = month(date, label = TRUE)) %>%
  ggplot() +
  geom_tile(aes(x = x, y=y, fill=.epred_sd)) +
  geom_sf(data = scale_72_clusters, colour="grey78", fill=NA) +
  #geom_sf(data = main_roads_cropped, colour="white") +
  scale_fill_viridis_c(option = "F") +
  facet_wrap(date~.) +
  labs(title = "Log(PM2.5) exposure",
       x="",
       y="",
       caption = "White boundaries are SCALE study clusters") +
  theme_ggdist() +
  theme(panel.background = element_rect(fill=NA, colour="grey78"),
        strip.background = element_rect(colour = "grey78"),
        axis.text.x = element_text(angle=90, hjust=1))

```

Sample coordinates, and plot over time

```{r, fig.width=12, fig.height=12}

#sample coordinate points
#will sort out the small number later
set.seed(123)
sampled_points_m2 <- aq_model_data %>%
  sample_n(50) %>%
  select(x, y, mean_temp_c, mean_current_humidity, pop_density_km2, building_coverage_pct, dist_to_road_m, grid_id, log_pm2_5)

#generate DOY 0–365 and calculate Fourier terms
doy_grid <- tibble(doy = seq(0,365, by=1)) %>%
  mutate(
    sin_doy1 = sin(2 * pi * doy / 365),
    cos_doy1 = cos(2 * pi * doy / 365),
    sin_doy2 = sin(4 * pi * doy / 365),
    cos_doy2 = cos(4 * pi * doy / 365),
    sin_doy3 = sin(6 * pi * doy / 365),
    cos_doy3 = cos(6 * pi * doy / 365)
  )

#expand grid across sampled locations
prediction_df_m2 <- sampled_points_m2 %>%
  crossing(doy_grid)

#add predictions
preds_m2 <- add_epred_draws(object = m2, newdata = prediction_df_m2, ndraws = 100)

#observed data for plotting
observed_data <- aq_model_data %>%
  select(mean_doy, log_pm2_5) %>%
  mutate(mean_doy = round(mean_doy))

#plot
preds_m2 %>%
  ungroup() %>%
  ggplot(aes(x=doy)) +
  stat_lineribbon(aes(y=.epred), .width=0.95) + 
  geom_jitter(data = observed_data, aes(x = mean_doy, y = log_pm2_5),
              color = "darkred", alpha = 0.6, width = 0.5, height = 0.0, size = 1.2) +
  scale_fill_brewer() +
  labs(title = "Model-estimated log(PM2.5) with empirical measurements",
       subtitle = "Model predictions with 95% CrI and observed data points",
       x = "Day of year",
       y = "log(PM2.5)",
       caption = "Modelled estimates restricted to within clusters") +
  theme_ggdist() +
  theme(panel.background = element_rect(colour = "grey78"),
        legend.position = "none")



```


Compare models, using LOO CV

```{r}

library(loo)

loo_m1 <- loo(m1)
loo_m2 <- loo(m2)

loo_compare(loo_m1, loo_m2)
```


TODO

 - OTHER OUTCOME MEASURES (PM1, PM10, NO)
 - CAN WE LINK TO THE STATIC AQ MONITORS (OUTDOORS)
 - PRIORS
 - OTHER COVARIATES
 - COMPARE MODELS WITH DIFFERENT BASIS FUNCTIONS FOR GP AND SPLINES
 - EXCEEDANCES
 - LINK TO MTB INFECTION PREVALENCE DATA, AND MODEL
