---
title: "Blantyre air quality modelling"
output:
  html_document:
    df_print: paged
---

Load required libraries

```{r}
library(tidyverse)
library(here)
library(brms)
library(tidybayes)
library(sf)
library(mgcv)
library(priorsense)
library(osmdata)
library(nngeo) 
```


### Air quality data

By summary, a random household survey was conducted in Blantyre Malawi during 2019-2020 (pre-COVID in Malawi) as part of a TB cluster randomised trial  pre-intervention prevalence survey (sumamrised here: https://europepmc.org/article/MED/37862284). Fieldworkers carried purple air monitors attached to their backpacks, with measurements set to read every 90 seconds. Not all fieldworkers had a monitor. Previously, Helen Savage cleaned these monitor data, and merged to household questionnaire data, filtering measurements by interview times. Therefore, measurements are indoor household measurements.

Cluster boundaries (72 in total) are based on Ministry of Health Community Health Worker catchment areas, refined in collaboration with researchers. These were the unit of randomisation for the SCALE cluster randomised trial.

All air quality measurements were then taken from households within the 72 SCALE cluster.

```{r}
load("input_data/aq_in.RData") #cleaned air-quality dataset
load("input_data/scale_72_clusters.rda") #SCALE clusters
```

Get data into correct shape and aggregate measurements per household. Note as purple air devices took measurements every 90 seconds, mostly housholds surveyed have multiple measurements. Here, for convenience, we take the mean of these measurement per household.

```{r, fig.width=12, fig.height=12}
aq_dat <- aq_in %>%
  select(datetime, h02cl_id, hh_id, hh_lat, hh_lon, temp_c, current_humidity, mean_pm_2_5, distance_km) %>%
 mutate(
   datetime = ymd_hms(datetime),
   hour = hour(datetime) + minute(datetime)/60,
   doy = yday(datetime),
   x = hh_lon,
   y = hh_lat) %>%
  group_by(hh_id) %>%
  reframe(h02cl_id = h02cl_id,
          x = x,
          y = y,
          mean_doy = mean(doy, na.rm=TRUE),
          mean_pm2_5 = mean(mean_pm_2_5), 
          mean_temp_c = mean(temp_c, na.rm=TRUE),
          mean_current_humidity = mean(current_humidity, na.rm=TRUE)) %>%
  distinct() %>%
  mutate(log_pm2_5 = log(mean_pm2_5))


#we will use Fourier series components with multiple harmonics to model day-of-the year effects
aq_dat <- aq_dat %>%
  mutate(
    sin_doy1 = sin(2 * pi * mean_doy / 365),
    cos_doy1 = cos(2 * pi * mean_doy / 365),
    sin_doy2 = sin(4 * pi * mean_doy / 365),
    cos_doy2 = cos(4 * pi * mean_doy / 365),
    sin_doy3 = sin(6 * pi * mean_doy / 365),
    cos_doy3 = cos(6 * pi * mean_doy / 365)
  )

#check some variables
aq_dat %>%
  select(hh_id, log_pm2_5, mean_temp_c, mean_current_humidity, mean_doy) %>%
  pivot_longer(cols = c(log_pm2_5, mean_temp_c, mean_current_humidity, mean_doy)) %>%
  ggplot() +
  geom_histogram(aes(x=value, fill=name)) +
  facet_wrap(name~., scales = "free_x")

#PM2.5 distribution by day of the year
aq_dat %>%
  ggplot() +
  geom_jitter(aes(x = mean_doy, y=log_pm2_5), colour="darkred", alpha=0.5) +
  theme_ggdist() +
  theme(panel.background = element_rect(colour="grey78"))
#Note that there was no data collection in April of either year


#plot the measurements on a map
aq_dat %>%
  ggplot() +
  geom_sf(data=scale_72_clusters, fill=NA) +
  geom_point(aes(x=x, y=y, colour=log_pm2_5), size=0.1)+
  scale_color_viridis_c() +
  theme_ggdist() +
  theme(panel.background = element_rect(colour="grey78")) 

#Note the warning about old coordinate reference system - we will fix this below.

```

### Covariates

Population counts and density



```{r, fig.width=12, fig.height=12}

#read in the .tiff image of population count from Worldpop
#https://hub.worldpop.org/geodata/summary?id=1560
mw_100m <- stars::read_stars("input_data/mwi_ppp_2020.tif")


#define buffer around
buffer_m <- 500

#approximate conversion factor (degrees per meter)
#adjust for latitude for longitude direction
lat_mean <- mean(aq_dat$y)
deg_per_m_lat <- 1 / 111320  # Rough conversion for latitude
deg_per_m_lon <- 1 / (111320 * cospi(lat_mean / 180))  # Adjust for latitude

#expandbounding box with buffer
bbox_buffered <- st_bbox(c(
  xmin = min(aq_dat$x) - buffer_m * deg_per_m_lon,
  xmax = max(aq_dat$x) + buffer_m * deg_per_m_lon,
  ymin = min(aq_dat$y) - buffer_m * deg_per_m_lat,
  ymax = max(aq_dat$y) + buffer_m * deg_per_m_lat
), crs = st_crs(mw_100m))

#crop worldpop raster to buffered bbox
mw_100m_cropped <- st_crop(mw_100m, bbox_buffered)

#plot
plot(mw_100m_cropped)


#there are a small number of cells in the centre that are NA
#I think these are the military radar on Mount Soche,
#so no expected population
#replace with 0
mw_100m_cropped[[1]][is.na(mw_100m_cropped[[1]])] <- 0

#did it work?
plot(mw_100m_cropped)

#xonvert to sf object
aq_sf <- st_as_sf(aq_dat, coords = c("x", "y"), crs = st_crs(mw_100m_cropped))

#merge with the population count data
aq_sf$pop_density <- stars::st_extract(mw_100m_cropped, aq_sf)[[1]]

#get population density
aq_model_data <- aq_sf %>%
  st_drop_geometry() %>%
  as_tibble() %>%
  mutate(pop_density_km2 = pop_density / 0.01) %>%
  left_join(aq_dat %>% select(hh_id, x, y))

#fix the coordinate crs of the scale clusters
scale_72_clusters <-  st_transform(scale_72_clusters, st_crs(mw_100m))

```

Building density (i.e. we get the building footprints, and calculate their combined footprint area per grid cell, then calculate what percentage of the grid cell is "building footprint")
https://sites.research.google/gr/open-buildings/#open-buildings-download 

Using Version 2 here - can update later to V3, with better precision, and year-specific data. (I think this is for 2020 currently)

(Buildings dataset is huge, so processing here, then importing. Preprocessing commented out.)

```{r, fig.width=12, fig.height=12}

#read in the data set
#is a stars dataset
buildings <- read_rds("input_data/blantyre_buff_buildings.rds")

#match the crs to the population data
buildings <-  st_transform(buildings, st_crs(mw_100m))

#convert stars object to polygons
mw_100m_grid_sf <- st_as_sf(mw_100m_cropped, as_points = FALSE, merge = FALSE)

#add grid cell ID for grouping
mw_100m_grid_sf <- mw_100m_grid_sf %>%
  mutate(grid_id = row_number())

#join building footprint data to the population data
buildings_with_grid <- st_join(buildings, mw_100m_grid_sf, left = FALSE)

#filter by confidence of building footprint, and by footprint size
#in Hannah R's previous experiements, this was a good compromise
#and matches well with home visits on the ground.
buildings_with_grid <- buildings_with_grid %>%
  filter(area_in_meters>20) %>%
  filter(area_in_meters<300) %>%
  filter(confidence>0.69)

#sum building footprint per grid cell
building_area_per_cell <- buildings_with_grid %>%
  group_by(grid_id) %>%
  summarise(total_building_area_m2 = sum(area_in_meters, na.rm = TRUE))


write_rds(building_area_per_cell, "input_data/building_area_per_cell.rds")

building_area_per_cell <- read_rds("input_data/building_area_per_cell.rds")

#cell area in m squared
cell_area_m2 <- 100 * 100

#compute percentage coverage
building_area_per_cell <- building_area_per_cell %>%
  mutate(building_coverage_pct = (total_building_area_m2 / cell_area_m2) * 100)

#join
mw_100m_grid_sf <- mw_100m_grid_sf %>%
  left_join(building_area_per_cell %>% st_drop_geometry(), by = "grid_id") %>%
  mutate(building_coverage_pct = replace_na(building_coverage_pct, 0))

#check by plotting
mw_100m_grid_sf %>%
  mutate(x = st_coordinates(st_centroid(.))[, 1],
         y = st_coordinates(st_centroid(.))[, 2]) %>%
  ggplot() +
  geom_tile(aes(x=x, y=y, fill=building_coverage_pct)) +
  scale_fill_viridis_c(option = "cividis") +
  theme_ggdist() +
  theme(panel.background = element_rect(colour = "grey78"))

#Look at building footprints in one cluster

cluster_buildings <- st_join(scale_72_clusters, buildings)

 cluster_buildings$footprints_geom <- st_as_sfc(cluster_buildings$footprints, crs = st_crs(cluster_buildings))
 st_geometry(cluster_buildings) <- "footprints_geom"

 cluster_buildings %>%
   filter(cluster =="c1") %>%
   ggplot() +
   geom_sf(fill = "mediumseagreen") +
   geom_sf(data = scale_72_clusters %>% filter(cluster == "c1"), fill=NA)


```

Distance to the nearest main road. I.e. because we think that PM2.5 exposure can come from road vehicles.

We use Open Street Map (OSM) data to get the main roads (there are few of these in Blantyre... but see the most traffic). Later could consider including some smaller road...

```{r, fig.width=12, fig.height=12}

#get the OSM features.
q <- opq(bbox = bbox_buffered) %>%
  add_osm_feature(key = 'highway', 
                  value = c('primary', 'secondary', 'tertiary', 
                            'trunk', 'motorway'))

#import
osm_roads <- osmdata_sf(q)

##get ready.
main_roads <- osm_roads$osm_lines
main_roads <- st_transform(main_roads, st_crs(mw_100m))

#plot roads
main_roads %>%
  ggplot() +
  geom_sf(data = scale_72_clusters, colour="darkred") +
  geom_sf() 

#crop to grid
main_roads_cropped <- st_crop(main_roads, bbox_buffered)

#plot again.
main_roads_cropped %>%
  ggplot() +
  geom_sf(data = scale_72_clusters, colour="darkred") +
  geom_sf() 

#get centroids of grid cells
grid_centroids <- st_centroid(mw_100m_grid_sf)

#use nearest neighbour search to get the distance from the grid cell centroid to the nearest main road.
nearest_roads <- st_nn(grid_centroids, main_roads_cropped, k = 1, returnDist = TRUE)
dist_to_road_m <- sapply(nearest_roads$dist, `[`, 1)

#same as before: add to grid
mw_100m_grid_sf$dist_to_road_m <- dist_to_road_m

#check plot
mw_100m_grid_sf %>%
  mutate(x = st_coordinates(st_centroid(.))[, 1],
         y = st_coordinates(st_centroid(.))[, 2]) %>%
  ggplot() +
  geom_tile(aes(x=x, y=y, fill=dist_to_road_m)) +
  scale_fill_viridis_c(option = "G", direction = -1) +
  theme(panel.background = element_rect(colour = "grey78")) +
  theme_ggdist() +
  theme(panel.background = element_rect(colour = "grey78"))



```


Join all together into the modelling dataset, and just last tidy-up.

```{r, fig.width=12, fig.height=12}

aq_model_data <- st_as_sf(aq_model_data, coords = c("x", "y"), crs = st_crs(mw_100m))

aq_model_data <- st_join(aq_model_data, mw_100m_grid_sf)

aq_model_data <- aq_model_data %>%
  mutate(x = st_coordinates(st_centroid(.))[, 1],
         y = st_coordinates(st_centroid(.))[, 2]) %>%
  st_drop_geometry() %>%
  mutate(building_coverage_pct = building_coverage_pct/100)

write_rds(aq_model_data, "input_data/aq_model_data.rds")
write_rds(mw_100m_cropped, "input_data/mw_100m_cropped.rds")
write_rds(mw_100m_grid_sf, "input_data/mw_100m_grid_sf.rds")
```
